### 2022-04-15 FALCON_description.tex
healing

To support large viscosities variations the penalty parameter is related to the effective elemental viscosity by means of a dimensionless coefficient, so that
$\lambda=\lambda_e(e)\eta_{eff}(e)$ \citep{Marotta2006,Dabrowski2008,Thieulot2014}. Benchmark of the falling block \citep{Gerya2003a,Gerya2010b,Thieulot2011}
is performed to verify that the code can correctly deal buoyancy driven flows with strong viscosity contrasts (Section \ref{sec:block}).

@@

@@

\section{The energy equation}\label{sec:en_eq}
The stabilisation of the advection term of the energy equation (Eq. \ref{eq:energy}) needed to avoid possible oscillations in the thermal solution in the case
when advection dominates over diffusion, is implemented by means of a streamline-upwind Petrov–Galerkin (SUPG) method, for which the advection term is modified
follows the discussion in \citet{Thieulot2011} and \citet{Thieulot2014} (see Section \ref{sec:advection} for benchmark). The total internal heat production
$H_{tot}$ in Eq. \ref{eq:energy} includes radiogenic heating $H_r$, shear heating $H_s$ and adiabatic heating $H_a$, where
\begin{equation}\label{eq:shear}
H_s=2\eta \dot{\bm{\epsilon}} : \dot{\bm{\epsilon}} = 2\eta(\dot{\epsilon}_{xx}^2+\dot{\epsilon}_{yy}^2+2\dot{\epsilon}_{xy}^2)
\end{equation}
and
\begin{equation}\label{eq:adiabatic}
H_a=T \alpha \frac{Dp}{Dt} \approx -\alpha T \rho g_y v
\end{equation}
The correctness of the implementation and the computation of shear and adiabatic heating is verified performing an exercise with an analytical solution
(Section \ref{sec:simple_shear}) and Exercise 9.4 in \citet{Gerya2010b} (Section \ref{sec:shear}). Furthermore, mantle convection and thin layer entrainment
benchmarks (Sections \ref{sec:mantle} and \ref{sec:thin}, respectively) are performed to verify that the code solve correctly Eqs. \ref{eq:momentum_penalty}
and \ref{eq:energy} in case of a temperature field characterised by an initial perturbation and with a temperature-dependent density. At last, the viscoplastic
mantle convection benchmark is performed to check also the viscoplasticity in case of a initial perturbed temperature field (Section \ref{sec:visco_mantle}).

@@

\subsection{Falling blocks}\label{sec:block}
This benchmark is described as proposed by \citet{Gerya2003a}, \citet{Gerya2010b} and \citet{Thieulot2011}. The domain is square with $L_x=L_y=$ \SI{500}{\km}
and the grid is composed by \num{50x50} elements with 25 markers in each element. The block is initially centred at ($x=$ \SI{250}{\km}; $y=$ \SI{400}{\km})
and has a size of \num{100x100} km (Fig. \ref{fig:block}a). The fluid surrounding the block has $\eta_f=$ \SI{e21}{\pascal\per\s} and $\rho_f=$
\SI{3200}{\kg\per\cubic\m}. The benchmark tests with different viscosities of the block, with $\eta_b$ from \SIrange{e15}{e27}{\pascal\per\s}. In each
experiment also the density $\rho_b$ of the block varies from \SIrange{3220}{9900}{\kg\per\cubic\m}. Velocity boundary conditions are set to free slip
conditions on all sides of the domain. Density distributions at $t=$ \SI{20}{\mega\year} for $\rho_b=$ \SI{3300}{\kg\per\cubic\m} and different $\eta_b$
are plotted in Fig. \ref{fig:block}b-f, showing that the code correctly preserve the block geometry in case of large viscosity contrast, with a block stiffer
than the surrounding fluid (Fig. \ref{fig:block}f).

The velocity in the centre of the falling block at $t=0$ is measured for all experiments and, since the velocity should increase with density contrast, the
quantity $\bm{v}/(\rho_b-\rho_f)$ is plotted as function of the viscosity contrast. All results perfectly match with those from \citet{Gerya2010b} and they
line up on a single curve, demonstrating that the code can correctly deal with large viscosity and density contrasts (Fig. \ref{fig:falling}). All data can be 
found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum_equation/Falling%20blocks}.

@@

\subsection{Shear and adiabatic heating}\label{sec:shear}
This problem is performed as presented in Exercise 9.4 in \citet{Gerya2010b}. Two materials are vertically separated in a rectangular domain with $L_x=$
\SI{1000}{\km}, $L_y=$ \SI{1500}{\km} and a grid resolution of \num{30x20} elements. Constant thermal coefficient expansion ($\alpha=$ \SI{3e-5}{\per\kelvin}),
temperature ($T=$ \SI{1300}{\kelvin}) and gravity acceleration ($g_y=$ \SI{-10}{\m\square\s}) are assumed in the whole domain. Fluid 1 (on the left side of the
domain) has $\rho_1=$ \SI{3200}{\kg\per\cubic\m} and $\eta_1=$ \SI{e20}{\pascal\s}; fluid 2 (on the right side of the domain) has $\rho_2=$
\SI{3300}{\kg\per\cubic\m} and $\eta_2=$ \SI{e22}{\pascal\s}. Velocity boundary conditions are set to free slip on all sides of the domain.

As shown in Fig. \ref{fig:energy}, both shear and adiabatic heating predicted by the code (first row in Fig. \ref{fig:energy}) well recreate the results
obtained by means of example ${Shear\_adiabatic\_heating.m}$ from \citet{Gerya2010b} (second row in Fig. \ref{fig:energy}). 
All data can be found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Energy_equation/Adiabatic%2BShear}.

\begin{figure}
\centering
\noindent\includegraphics[width=400px]{./Figures/Energy.pdf}
\caption{Comparison between shear (first column), adiabatic (second column) and total (third column) energy predicted by the code (panels a, b and c) and those
created using example ${Shear\_adiabatic\_heating.m}$ from Exercise 9.4 in \citet{Gerya2010b} (panels d, e and f).}
\label{fig:energy}
\end{figure}

>>>>

To support large viscosities variations the penalty parameter is related to the effective elemental viscosity by means of a dimensionless coefficient, so that
$\lambda=\lambda_e(e)\eta_{eff}(e)$ \citep{Marotta2006,Dabrowski2008,Thieulot2014}. Benchmark of the falling block \citep{Gerya2003a,Gerya2010,Thieulot2011}
is performed to verify that the code can correctly deal buoyancy driven flows with strong viscosity contrasts (Section \ref{sec:block}).

@@

\section{Healing and weakening}
The temporal evolution of the accumulated strain $\epsilon$ has been implemented as in \citet{Fuchs2019,Fuchs2021}, as following
\begin{equation}\label{eq:healing}
  \frac{d\epsilon}{dt}=I_2 - \epsilon H_{rate}(T)
\end{equation}
where the first term on the right-hand side is a source term given by the second invariant of the strain rate $I_2$ and the second term a
temperature-dependent healing factor $H_{rate}$, calculated as
\begin{equation}\label{eq:healing_rate}
  H_{rate}(T)=B \exp \left[-\frac{\mu}{2}\left(\frac{1}{T+1}-\frac{1}{2}\right)\right]
\end{equation}
where $B$ is a constant describing the time scale of healing, while $\mu$ and $T$ are the non-dimensional temperature activation and temperature,
respectively \citep{Fuchs2019,Fuchs2021}.

@@

\section{The energy equation}\label{sec:en_eq}
The stabilisation of the advection term of the energy equation (Eq. \ref{eq:energy}) needed to avoid possible oscillations in the thermal solution in the case
when advection dominates over diffusion, is implemented by means of a streamline-upwind Petrov–Galerkin (SUPG) method, for which the advection term is modified
follows the discussion in \citet{Thieulot2011} and \citet{Thieulot2014} (see Section \ref{sec:advection} for benchmark). The total internal heat production
$H_{tot}$ in Eq. \ref{eq:energy} includes radiogenic heating $H_r$, shear heating $H_s$ and adiabatic heating $H_a$, where
\begin{equation}\label{eq:shear}
H_s=2\eta \dot{\bm{\epsilon}} : \dot{\bm{\epsilon}} = 2\eta(\dot{\epsilon}_{xx}^2+\dot{\epsilon}_{yy}^2+2\dot{\epsilon}_{xy}^2)
\end{equation}
and
\begin{equation}\label{eq:adiabatic}
H_a=T \alpha \frac{Dp}{Dt} \approx -\alpha T \rho g_y v
\end{equation}
The correctness of the implementation and the computation of shear and adiabatic heating is verified performing an exercise with an analytical solution
(Section \ref{sec:simple_shear}) and Exercise 9.4 in \citet{Gerya2010} (Section \ref{sec:shear}). Furthermore, mantle convection and thin layer entrainment
benchmarks (Sections \ref{sec:mantle} and \ref{sec:thin}, respectively) are performed to verify that the code solve correctly Eqs. \ref{eq:momentum_penalty}
and \ref{eq:energy} in case of a temperature field characterised by an initial perturbation and with a temperature-dependent density. At last, the viscoplastic
mantle convection benchmark is performed to check also the viscoplasticity in case of a initial perturbed temperature field (Section \ref{sec:visco_mantle}).

@@

\subsection{Falling blocks}\label{sec:block}
This benchmark is described as proposed by \citet{Gerya2003a}, \citet{Gerya2010} and \citet{Thieulot2011}. The domain is square with $L_x=L_y=$ \SI{500}{\km}
and the grid is composed by \num{50x50} elements with 25 markers in each element. The block is initially centred at ($x=$ \SI{250}{\km}; $y=$ \SI{400}{\km})
and has a size of \num{100x100} km (Fig. \ref{fig:block}a). The fluid surrounding the block has $\eta_f=$ \SI{e21}{\pascal\per\s} and $\rho_f=$
\SI{3200}{\kg\per\cubic\m}. The benchmark tests with different viscosities of the block, with $\eta_b$ from \SIrange{e15}{e27}{\pascal\per\s}. In each
experiment also the density $\rho_b$ of the block varies from \SIrange{3220}{9900}{\kg\per\cubic\m}. Velocity boundary conditions are set to free slip
conditions on all sides of the domain. Density distributions at $t=$ \SI{20}{\mega\year} for $\rho_b=$ \SI{3300}{\kg\per\cubic\m} and different $\eta_b$
are plotted in Fig. \ref{fig:block}b-f, showing that the code correctly preserve the block geometry in case of large viscosity contrast, with a block stiffer
than the surrounding fluid (Fig. \ref{fig:block}f).

The velocity in the centre of the falling block at $t=0$ is measured for all experiments and, since the velocity should increase with density contrast, the
quantity $\bm{v}/(\rho_b-\rho_f)$ is plotted as function of the viscosity contrast. All results perfectly match with those from \citet{Gerya2010} and they
line up on a single curve, demonstrating that the code can correctly deal with large viscosity and density contrasts (Fig. \ref{fig:falling}). All data can be 
found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum_equation/Falling%20blocks}.

@@

\subsection{Shear and adiabatic heating}\label{sec:shear}
This problem is performed as presented in Exercise 9.4 in \citet{Gerya2010}. Two materials are vertically separated in a rectangular domain with $L_x=$
\SI{1000}{\km}, $L_y=$ \SI{1500}{\km} and a grid resolution of \num{30x20} elements. Constant thermal coefficient expansion ($\alpha=$ \SI{3e-5}{\per\kelvin}),
temperature ($T=$ \SI{1300}{\kelvin}) and gravity acceleration ($g_y=$ \SI{-10}{\m\square\s}) are assumed in the whole domain. Fluid 1 (on the left side of the
domain) has $\rho_1=$ \SI{3200}{\kg\per\cubic\m} and $\eta_1=$ \SI{e20}{\pascal\s}; fluid 2 (on the right side of the domain) has $\rho_2=$
\SI{3300}{\kg\per\cubic\m} and $\eta_2=$ \SI{e22}{\pascal\s}. Velocity boundary conditions are set to free slip on all sides of the domain.

As shown in Fig. \ref{fig:energy}, both shear and adiabatic heating predicted by the code (first row in Fig. \ref{fig:energy}) well recreate the results
obtained by means of example ${Shear\_adiabatic\_heating.m}$ from \citet{Gerya2010} (second row in Fig. \ref{fig:energy}). 
All data can be found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Energy_equation/Adiabatic%2BShear}.

\begin{figure}
\centering
\noindent\includegraphics[width=400px]{./Figures/Energy.pdf}
\caption{Comparison between shear (first column), adiabatic (second column) and total (third column) energy predicted by the code (panels a, b and c) and those
created using example ${Shear\_adiabatic\_heating.m}$ from Exercise 9.4 in \citet{Gerya2010} (panels d, e and f).}
\label{fig:energy}
\end{figure}
